### 2024-06-26 parts/intro.tex
up to line 150

\paragraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
the multiple characterizations of \kl{$\Nat$-polyregular function} (resp.
$\Rel$-polyregular functions), show that one can decide if a function $f \in
\ZPoly$ is \kl{commutative} (\cref{decidable-commutative-poly:lemma}), and give
the basic definitions used on multivariate polynomials. In
\cref{polynomials:sec}, we focus on \emph{polynomials}, by introducing the
\cref{karh:thm} of \cite{KARH77}, providing a counter-example in
\cref{thm:counter-example}, and an alternative characterization in
\cref{corrected-version:thm}. In \cref{beyond-polynomials:sec}, we leverage the
results over polynomials to decide $\NPoly$ inside $\ZPoly$ in the
\kl{commutative} case (\cref{decidable-n-poly:thm}), and prove that
\cref{zsf-nsf:conjecture}
\cite[Conjecture 7.61]{DOUE23} 
holds for \kl{commutative} functions
(\cref{zsf-npoly-nsf:thm}). Finally, in \cref{beyond-commutative:sec}, we
introduce the notion of \kl{residual transducer}
(\cref{residual-transducer:def}), and prove in \cref{non-commutative-npoly:thm}
that it characterizes $\NPoly$ inside $\ZPoly$ through a notion of
\kl{well-quasi-ordered} collection of residuals.

>>>>

\paragraph*{Outline of the paper.} In \cref{preliminaries:sec}, we introduce
the multiple characterizations of \kl{$\Nat$-polyregular functions} (resp.
$\Rel$-polyregular functions), show that one can decide if a function $f \in
\ZPoly$ is \kl{commutative} (\cref{decidable-commutative-poly:lemma}), and give
the basic definitions used on multivariate polynomials. In
\cref{polynomials:sec}, we focus on \emph{polynomials}, by introducing the
\cref{karh:thm} of \cite{KARH77}, providing a counter-example in
\cref{thm:counter-example}, and an alternative characterization in
\cref{corrected-version:thm}. In \cref{beyond-polynomials:sec}, we leverage the
results over polynomials to decide $\NPoly$ inside $\ZPoly$ in the
\kl{commutative} case (\cref{decidable-n-poly:thm}), and prove that
\cref{zsf-nsf:conjecture}
\cite[Conjecture 7.61]{DOUE23} 
holds for \kl{commutative} functions
(\cref{zsf-npoly-nsf:thm}). Finally, in \cref{beyond-commutative:sec}, we
introduce the notion of \kl{residual transducer}
(\cref{residual-transducer:def}), and prove in \cref{non-commutative-npoly:thm}
that it characterizes $\NPoly$ inside $\ZPoly$ through a notion of
\kl{well-quasi-ordered} collection of residuals.

### 2024-06-26 parts/prelim.tex
up to line 150

\AP In the rest of this paper, $\Rel$ (resp. $\Nat$) denotes the set of
integers (resp. non-negative integers). The capital letters $\Sigma,\Gamma$
denotes fixed alphabets, i.e. a finite set of letters, and $\Sigma^*, \Gamma^*$
(resp. $\Sigma^+, \Gamma^+$) are the set of words (resp. non-empty words) over
$\Sigma, \Gamma$. The empty word is written $\varepsilon \in \Sigma^*$. When $w
\in \Sigma^*$ and $a \in \Sigma$, we let $\card{w} \in \Nat$ be the length of
$w$, and $\card[a]{w}$ be the number of occurrences of $a$ in $w$. We assume
that the reader is familiar with the basics of automata theory, in particular
the notions of monoid morphisms, idempotents in monoids, monadic second-order
($\MSO$) logic and first-order ($\FO$) logic over finite words (see e.g.
\cite{THOM97}). As aperiodicity will be a central notion of this paper, let us
recall that a monoid $M$ is \intro(monoid){aperiodic} whenever for all $x \in
M$, there exists $n \in \Nat$ such that $x^{n+1} = x^n$.

\AP We also assume the reader to be familiar with \kl{rational series} and
weighted automata, and refer to \cite{BERE10} for a comprehensive introduction.
Let us recall that one of the equivalent definitions of a $K$-rational series
(where $K$ is a ring) is given by the functions from $\Sigma^*$ to $K$ computed
by weighted automata, which are non-deterministic finite automata with
transitions labelled by elements of $K$, and whose output on a given word is
the sum over all accepting paths of the products of the weights along the path.

\AP We use the notation $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the map
that counts occurrences of every letter in the input word (that is, computes
the Parikh vector) namely: $ \commute[w] \defined (a \mapsto \card[a]{w})$.
Given a set $X$, a function $f \colon \Sigma^* \to X$ is \intro{commutative}
whenever for all $u \in \Sigma^*$, for all permutations $\sigma$ of $\set{1,
\dots, \card{w}}$, $f(\sigma(u)) = f(u)$. Equivalently, it is
\reintro{commutative} whenever there exists a map $g \colon \Nat^\Sigma \to X$
such that $g \circ \commute = f$. A function $g \colon \Nat^n \to X$ is
\intro{represented} by a function $f \colon \Sigma^* \to X$ if there exists a
map $\eta \colon \set{1, \dots, n} \to \Sigma$ such that $g \circ (v \colon
\Nat^\Sigma \mapsto \seqof{v(\eta(k))}{1 \leq k \leq n} \colon \Nat^n) \circ
\commute = f$. This notion will be useful to formally state that a polynomial
``is'' a \kl{polyregular function}.

@@

\begin{definition}
    Let $\mathcal{H}$ be a family of functions
    from $\Sigma^*$ to $\Rel$.
    An \intro{$\mathcal{H}$-transducer} $\aTransd$ is
    a tuple $(Q, q_0, \delta, \lambda, F)$ where
         $Q$ is a finite set of states,
         $q_0 \in Q$ is an initial state,
         $\delta \colon Q \times \Sigma \to Q$
            is a deterministic transition function,
         $\lambda \colon Q \times \Sigma \to \mathcal{H}$
            is a correction function,
         and $F \colon Q \to \Rel$ is an output function.
\end{definition}

\begin{definition}
    Let $\aTransd \defined (Q,q_0, \delta, \lambda, F)$ be an \kl{$\mathcal{H}$-transducer}.
    The function
    $\aTransd \colon Q \times \Sigma^* \to \Rel$
    is defined inductively as follows:
        $\aTransd(q, \varepsilon) \defined F(q)$, and
        $\aTransd(q, a u) \defined \aTransd(\delta(q,a), u)
            + \lambda(q,a)(u)$.
    The function \intro{computed} by a transducer $\aTransd$
    is $w \mapsto \aTransd(q_0, w)$.
\end{definition}

@@

\begin{example}
    \label{non-canonical-transd:ex}
    Let $\BadExOk \colon \set{a}^* \to \Nat$
    be defined as
    $\BadExOk(\varepsilon) = 1$,
    $\BadExOk(a) = 0$,
    and $\BadExOk(a^n) = n - 1$ for all $n \geq 2$.
    The function $\BadExOk$ is \kl{computed} by the two \kl{$\NPoly[0]$-transducers}
    depicted in \cref{non-canonical-transd:fig}.
\end{example}

@@

            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid 1 + \ind{\card{w} \geq 1}$} (A);
        \end{tikzpicture}
        \caption{This automaton has a \kl{counter}.}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\linewidth}
        \begin{tikzpicture}[
            etat/.style={minimum size=2em}
            ]
            \node[etat,state,initial,
                accepting by arrow,
                accepting text={$1$},
                accepting where=below,
                ] (A) at (0,0) {$\varepsilon$};
            \node[etat,state,
                accepting by arrow,
                accepting text={$0$},
                accepting where=below,
                ] (B) at (2,0) {$a$};

@@

            \draw[->] (B) edge[loop right] node[midway, right]
                {$a \mid 1 + \ind{\card{w} = 0}$} (B);
        \end{tikzpicture}
        \caption{This automaton is \kl{counter-free}.}
    \end{subfigure}
    \caption{Two \kl{$\NPoly[0]$-transducers}
    \kl{computing} the function $\BadExOk$ of \cref{non-canonical-transd:ex}.}
    \label{non-canonical-transd:fig}
\end{figure}

>>>>

\AP In the rest of this paper, we use the symbol $\Rel$ to denote the set of
integers, $\Rat$ to denote the set of rational numbers, and $\Nat$ to denote
the set of non-negative integers. The capital letters $\Sigma,\Gamma$ denote
fixed alphabets, i.e. finite set of letters, and $\Sigma^*, \Gamma^*$ (resp.
$\Sigma^+, \Gamma^+$) are the set of words (resp. non-empty words) over
$\Sigma, \Gamma$. The empty word is written $\varepsilon \in \Sigma^*$. When $w
\in \Sigma^*$ and $a \in \Sigma$, we let $\card{w} \in \Nat$ be the length of
$w$, and $\card[a]{w}$ be the number of occurrences of $a$ in $w$.

\AP
We assume that the reader is familiar with the basics of automata theory, in
particular the notions of monoid morphisms, idempotents in monoids, monadic
second-order ($\MSO$) logic and first-order ($\FO$) logic over finite words
(see e.g. \cite{THOM97}). As aperiodicity will be a central notion of this
paper, let us recall that a monoid $M$ is \intro(monoid){aperiodic} whenever
for all $x \in M$, there exists $n \in \Nat$ such that $x^{n+1} = x^n$.

\AP Let us briefly recall that one of the equivalent definitions of a
\intro{$K$-rational series} (where $K$ is a ring), and refer to \cite{BERE10}
for a comprehensive introduction. The \kl{$K$-rational series} from $\Sigma^*$
to $K$ are the functions computed by \intro{$K$-weighted automata} which are
non-deterministic finite automata with transitions labelled by elements of $K$,
and whose output on a given word is the sum over all accepting paths of the
products of the weights along the path. For instance, the function $w \mapsto
2^{\card{w}}$ is computed by an \kl{$\Nat$-weighted automaton} with a single
state, and a loop labelled by $2$, and is therefore an \kl{$\Nat$-rational
series}.

\AP We use the notation $\commute \colon \Sigma^* \to \Nat^\Sigma$ for the map
that counts occurrences of every letter in the input word (that is, computes
the Parikh vector) namely: $ \commute[w] \defined \seqof{a \mapsto
\card[a]{w}}{a \in \Sigma}$. Given a set $X$, a function $f \colon \Sigma^* \to
X$ is \intro{commutative} whenever for all $u \in \Sigma^*$, for all
permutations $\sigma$ of $\set{1, \dots, \card{w}}$, $f(\sigma(u)) = f(u)$.
Equivalently, it is \reintro{commutative} whenever there exists a map $g \colon
\Nat^\Sigma \to X$ such that $g \circ \commute = f$.

\AP Let $n \in \Nat$, and let $\Sigma$ be a finite alphabet. Given a function
$\eta \colon \set{1, \dots, n} \to \Sigma$, we define the $\eta^\dagger \colon
\Nat^n \to \Sigma^*$ as $\eta^\dagger(\vec{x}) \defined \eta(1)^{x_1} \dots
\eta(n)^{x_n}$. A function $f \colon \Nat^n \to X$ is \intro{represented} by a
\kl{commutative} function $g \colon \Sigma^* \to X$ if there exists a map $\eta
\colon \set{1, \dots, n} \to \Sigma$ such that $g \circ \eta^\dagger = f$. This
notion will be useful to formally state that a polynomial ``is'' a
\kl{commutative} \kl{polyregular function}. For instance, the polynomial
function $P(X,Y) = X \times Y$ is \kl{represented} by the \kl{commutative}
function $g \colon \set{a,b}^* \to \Rel$ defined by $g(w) \defined \card[a]{w}
\times \card[b]{w}$.

@@

\begin{definition}
    Let $\mathcal{H}$ be a family of functions
    from $\Sigma^*$ to $\Rel$.
    An \intro{$\mathcal{H}$-transducer} $\aTransd$ is
    a tuple $(Q, q_0, \delta, \lambda, F)$ where
         $Q$ is a finite set of states,
         $q_0 \in Q$ is called the initial state of the transducer,
         $\delta \colon Q \times \Sigma \to Q$
            is called the deterministic transition function of the transducer,
         $\lambda \colon Q \times \Sigma \to \mathcal{H}$
            is called the correction function of the transducer,
         and $F \colon Q \to \Rel$ is called the output function of the transducer.
\end{definition}

Based on this syntax, let us now define the semantics of a transducer, in terms
of the function from $\Sigma^*$ to $\Rel$ that it computes. The intuition is
that it suffices to apply the correction function $\lambda$ on every transition
of the underlying automaton, sum the results, and add to this number the output
associated to the final state of the computation.

\begin{definition}
    \label{transducer-sem:def}
    Let $\aTransd \defined (Q,q_0, \delta, \lambda, F)$ be an \kl{$\mathcal{H}$-transducer}.
    The function
    $\aTransd \colon Q \times \Sigma^* \to \Rel$
    is defined inductively as follows:
        $\aTransd(q, \varepsilon) \defined F(q)$, and
        $\aTransd(q, a u) \defined \aTransd(\delta(q,a), u)
            + \lambda(q,a)(u)$.
    The function \intro{computed} by a transducer $\aTransd$
    is $w \mapsto \aTransd(q_0, w)$.
\end{definition}

@@

\begin{example}
    \label{non-canonical-transd:ex}
    Let $\BadExOk \colon \set{a}^* \to \Nat$
    be defined as
    $\BadExOk(\varepsilon) = 1$,
    $\BadExOk(a) = 0$,
    and $\BadExOk(a^n) = n - 1$ for all $n \geq 2$.
    The function $\BadExOk$ is \kl{computed} by the two \kl{$\NPoly[0]$-transducers}
    depicted in \cref{non-canonical-transd:fig}.
    As a consequence, $f \in \NPoly[1]$.
    Furthermore, because the automaton of 
        \cref{non-canonical-transd:fig:counter-free}
        has no \kl{counter},
    and the correction function only uses functions in $\NSF[0]$,
    $f \in \NSF[1]$.
\end{example}

@@

            \draw[->] (B) to[bend right=45] node[midway, above] {$a \mid \ind{\card{w} > 1} + \ind{\card{w} \geq 1}$} (A);
        \end{tikzpicture}
        \caption{This automaton has a \kl{counter}.}
    \end{subfigure}
    \begin{subfigure}[b]{0.49\linewidth}
        \begin{tikzpicture}[
            etat/.style={minimum size=2em}
            ]
            \node[etat,state,initial,
                accepting by arrow,
                accepting text={$1$},
                accepting where=below,
                ] (A) at (0,0) {$\varepsilon$};
            \node[etat,state,
                accepting by arrow,
                accepting text={$0$},
                accepting where=below,
                ] (B) at (2,0) {$a$};

@@

            \draw[->] (B) edge[loop right] node[midway, right]
                {$a \mid 1 + \ind{\card{w} = 0}$} (B);
        \end{tikzpicture}
        \caption{This automaton is \kl{counter-free}.}
        \label{non-canonical-transd:fig:counter-free}
    \end{subfigure}
    \caption{Two \kl{$\NPoly[0]$-transducers}
    \kl{computing} the function $\BadExOk$ of \cref{non-canonical-transd:ex}.}
    \label{non-canonical-transd:fig}
\end{figure}
